---
title: "Figure 3  Analyses of MIC and LEC scores"
author: "Douglas Yu"
date: "05 June 2018"
output: html_document
---
## Packages
```{r, message=FALSE}
# rm(list=ls())
library(tidyverse)
library(readxl)
library(cowplot)
library(knitr)
library(ggbeeswarm)
library(RColorBrewer)
library(ggthemes)
library(envDocument)
library(broom)
env_doc("table", git = FALSE)
```

```{r}
datafile <- "Figure_3_MIC_LEC_analyses_DataSheet_20180626.xlsx"
datasheet <- "Sheet1"
```

## Dataset 1: MIC and LEC scores
```{r Streptomyces dataset}
assay1 <- read_excel(datafile, datasheet, na = "NA")
assay1 <- assay1 %>% dplyr::select(Strain, Strep_Soft, Antibiotic, MIC_score, LEC_score, Notes)
assay1 <- assay1 %>% dplyr::filter(is.na(Notes) | Notes != "Contamination") %>% dplyr::select(-Notes)
assay1$Strain <- as_factor(assay1$Strain)
assay1$Strep_Soft <- as_factor(assay1$Strep_Soft)
assay1$Antibiotic <- as_factor(assay1$Antibiotic)
tbl_df(assay1) # should have 235 rows and 5 cols

xtabs(~ Strain + Strep_Soft, assay1) # to see experimental design
```


```{r stripchart of results}
colorvec <- brewer.pal(3,"RdBu")
colorvec <- c(colorvec[3], colorvec[1], colorvec[2])

mic <- ggplot(assay1, aes(x = Strep_Soft, y = MIC_score))
lec <- ggplot(assay1, aes(x = Strep_Soft, y = LEC_score)) 

(lecplot <- lec 
	+ geom_boxplot(size = .5, outlier.shape = NA, width = 0.3, aes(fill = assay1$Strep_Soft), notch = TRUE) 
	# + geom_jitter(width = 0.05, cex = .8)
    + geom_quasirandom()
	+ scale_shape_manual(values = c(16, 2)) 
	+ labs(y = expression("Lowest Inhibitory Concentration (LEC) score")) 
	+ scale_fill_manual(values = colorvec) 
    + theme(axis.title.y = element_text(size = 14), 
  			axis.title.x = element_blank(), 
  			legend.position='none', 
  			panel.background = element_rect(fill = "white"), 
  			panel.grid.major.y = element_line(colour = "grey", size = 0.5),
  			panel.grid.minor.y = element_line(colour = "grey", size = 0.1),
  			panel.border = element_rect(colour = "black", fill = NA)
  			)
    )

(micplot <- mic 
	+ geom_boxplot(size = .5, outlier.shape = NA, width = 0.3, aes(fill = assay1$Strep_Soft), notch = TRUE) 
	# + geom_jitter(width = 0.05, cex = .8)
    + geom_quasirandom()
	+ scale_shape_manual(values = c(16, 2)) 
	+ labs(y = expression("Minimum Inhibitory Concentration (MIC) score")) 
	+ scale_fill_manual(values = colorvec) 
    + theme(axis.title.y = element_text(size = 14), 
  			axis.title.x = element_blank(), 
  			legend.position='none', 
  			panel.background = element_rect(fill = "white"), 
  			panel.grid.major.y = element_line(colour = "grey", size = 0.5),
  			panel.grid.minor.y = element_line(colour = "grey", size = 0.1),
  			panel.border = element_rect(colour = "black", fill = NA)
  			)
    )

plot_grid(lecplot, micplot, labels = c("A", "B"))
```

```{r statistics}
levels(assay1$Strep_Soft)
median(assay1$MIC_score[assay1$Strep_Soft == "Antibiotic-producers"])
median(assay1$MIC_score[assay1$Strep_Soft == "Non-producers"])
median(assay1$MIC_score[assay1$Strep_Soft == "Resident Non-producers"])

median(assay1$LEC_score[assay1$Strep_Soft == "Antibiotic-producers"])
median(assay1$LEC_score[assay1$Strep_Soft == "Non-producers"])
median(assay1$LEC_score[assay1$Strep_Soft == "Resident Non-producers"])
```

```{r compare producers to nonproducers}
assay2 <- assay1 %>% dplyr::filter(Strep_Soft != "Resident Non-producers")

# summarise by Strain
assay2 <- assay2 %>% group_by(Strain) %>% 
    summarise(Strep_Soft = first(Strep_Soft), 
                 mean_MIC = mean(MIC_score),
                 mean_LEC = mean(LEC_score)
    )

assay2$Strep_Soft <- fct_drop(assay2$Strep_Soft) # drop unused factor level
levels(assay2$Strep_Soft) # confirm

mod1 <- wilcox.test(assay2$mean_MIC ~ assay2$Strep_Soft)
(mod1tidy <- tidy(mod1))
mod2 <- wilcox.test(assay2$mean_LEC ~ assay2$Strep_Soft)
(mod2tidy <- tidy(mod2))
pvalues <- c(mod1tidy$p.value, mod2tidy$p.value)
p.adjust(pvalues, "fdr")
```

```{r compare producers to resident nonproducers}
assay3 <- assay1 %>% dplyr::filter(Strep_Soft != "Non-producers")

# summarise by Strain
assay3 <- assay3 %>% group_by(Strain) %>% 
    summarise(Strep_Soft = first(Strep_Soft), 
                 mean_MIC = mean(MIC_score),
                 mean_LEC = mean(LEC_score)
    )

assay3$Strep_Soft <- fct_drop(assay3$Strep_Soft) # drop unused factor level
levels(assay3$Strep_Soft) # confirm

mod1 <- wilcox.test(assay3$mean_MIC ~ assay3$Strep_Soft)
(mod1tidy <- tidy(mod1))
mod2 <- wilcox.test(assay3$mean_LEC ~ assay3$Strep_Soft)
(mod2tidy <- tidy(mod2))
pvalues <- c(mod1tidy$p.value, mod2tidy$p.value)
p.adjust(pvalues, "fdr")
```

To read these results:

1. Antibiotic-producers have higher MIC and LEC scores than do Non-producers, especially for LEC scores, which is the minimum concentration needed to observe an inhibitory effect.

2. We conservatively average the mean MIC and LEC scores over all antibiotics for each Antibiotic-producer and Non-producer strain, reducing the dataset from n = 155 (assay1) to n = 20 (assay2)

3. We then apply a Wilcoxon 2-sided test of equality of means.  Mean MIC scores and LEC scores are significantly higher for Antibiotic-producers.  These differences hold up after adjusting for two tests (p.adjust, method = false discovery rate).

4. Although it's not really necessary, for completeness, I test for mean differences between Antibiotic-producers and Resident Non-producers (assay3), and as is obvious from the boxplot, there is no difference. 

