---
title: "Analysis of microbiome competition experiment - Figure 2"
author: "Douglas Yu"
date: "05 June 2018"
output: html_document
---

## Housekeeping
```{r, message=FALSE}
# rm(list=ls())
library(lme4)
library(tidyverse)
library(readxl)
library(forcats)
library(sjPlot)
library(RColorBrewer)
library(envDocument)
env_doc("table", git = FALSE)
```

Variables
```{r}
datafile <- "Figure_2_Competition_analyses_DataSheet_20180301.xlsx"
```

Import data and format
```{r Import data and format}
assay <- read_excel(file.path(datafile), sheet = "Sheet1", na = "NA", col_types = c("text", "text", "text", "text"))
tbl_df(assay) # should have 250 rows and 4 cols
assay$Strep_strain <- as_factor(assay$Strep_strain) # Streptomyces strain
assay$Softbug_strain <- as_factor(assay$Softbug_strain) # Softbug is our slang term for non-antibiotic-producer bacterial strains
assay <- assay %>% dplyr::filter((Strep_outcome != "Ambiguous")) # remove one outcome that could not be scored non-ambiguously
assay$medium <- as_factor(assay$medium)
assay$Strep_outcome <- as_factor(assay$Strep_outcome)

assay$Strep_outcome <- fct_collapse(assay$Strep_outcome,
  Draw = c("Draws"),
  Win = c("Wins"),
  Loss = c("Loses")
)

levels(assay$Strep_outcome)

assay$medium2 <- fct_collapse(assay$medium, 
        Ps1_infused_media = c("Ae707_Ps1"),
		Ps2_infused_media = c("Ae717_Ps2"),
		Control = c("Ctrl")
)

levels(assay$medium2)

assay <- assay %>% dplyr::select(medium, medium2, Strep_strain, Softbug_strain, Strep_outcome) %>% arrange(medium, Strep_outcome)

assay # should have 249 rows and 5 cols

# create numeric versions of the competition outcomes
assay$Strep_outcomeLW <- NA
assay$Strep_outcomeLW[assay$Strep_outcome=="Win"] <- 1
assay$Strep_outcomeLW[assay$Strep_outcome=="Loss"] <- 0
assay$Strep_outcomeLW[assay$Strep_outcome=="Draw"] <- NA
assay$Strep_outcomeLW <- as.numeric(assay$Strep_outcomeLW)
```

Separate into the two experiments
```{r}
assay8 <- assay %>% filter(Strep_strain == "Strep_8")
assay8$medium <- assay8$medium %>% fct_drop()
assay8$medium2 <- assay8$medium2 %>% fct_drop()
assay8$Strep_strain <- assay8$Strep_strain %>% fct_drop()

xtabs(~ Softbug_strain + medium, assay8)

assay2 <- assay %>% filter(Strep_strain == "Strep_2")
assay2$medium <- assay2$medium %>% fct_drop()
assay2$medium2 <- assay2$medium2 %>% fct_drop()
assay2$Strep_strain <- assay2$Strep_strain %>% fct_drop()

xtabs(~ Softbug_strain + medium, assay2)

```

Summary barplots
```{r}
colorvec <- brewer.pal(3,"RdBu")
colorvec <- c(colorvec[1], colorvec[3], colorvec[2])

assay8_table <-  with(assay8, table(Strep_outcome, medium2))
assay8_table

assay2_table <-  with(assay2, table(Strep_outcome, medium2))
assay2_table

# formatted for publication
par(mfrow=c(1,2))
barplot(assay8_table, beside=T, main = expression(paste("Competitive outcome for ", italic('Streptomyces'), " strain 8")), names.arg = c("Ps1-infused media", "Control media"), ylab="Frequency", col=colorvec, ylim = c(0, 80))
legend("topleft",c("Win", "Loss", "Draw"), fill=colorvec, horiz = TRUE)
barplot(assay2_table, beside=T, main = expression(paste("Competitive outcome for ", italic('Streptomyces'), " strain 2")), names.arg = c("Ps2-infused media", "Control media"), col=colorvec, ylim = c(0, 80))
legend("topleft",c("Win", "Loss", "Draw"), fill=colorvec, horiz = TRUE)
par(mfrow=c(1,1))
```

Remove the Draw outcomes for downstream analysis (and new barplots)
```{r}
assay2.2 <- assay2 %>% dplyr::filter(Strep_outcome != "Draw")
assay2.2$Strep_outcome <- assay2.2$Strep_outcome %>% fct_drop()

assay8.2 <- assay8 %>% dplyr::filter(Strep_outcome != "Draw")
assay8.2$Strep_outcome <- assay8.2$Strep_outcome %>% fct_drop()

assay2.2_table <-  with(assay2.2, table(Strep_outcome, medium2))
assay2.2_table

assay8.2_table <-  with(assay8.2, table(Strep_outcome, medium2))
assay8.2_table

colorvec2 <- c(colorvec[1], colorvec[2])

par(mfrow=c(1,2))
barplot(assay8.2_table, beside=T, main="Strepto Strain 8; Ae707_Ps1 medium", ylab="Frequency", xlab = "Medium",col=colorvec2)
legend("topleft",c("win", "loss"), fill=colorvec2)
barplot(assay2.2_table, beside=T, main="Strepto Strain 2; Ae717_Ps2 medium", ylab="Frequency", xlab = "Medium", col=colorvec2, ylim = c(0, 80))
legend("topright",c("win", "loss"),fill=colorvec2)
par(mfrow=c(1,1))
```

# Start of analysis.  First analyse naively without random factor grouping the non-producer (softbug) strains

Highly significant interaction effect:  growth medium has a strong effect on competitive outcomes, for both Steptomyces strains
```{r}
model2.2<-chisq.test(assay2.2_table, correct = F) # basic chi-square test
model2.2$expected
model2.2

model8.2<-chisq.test(assay8.2_table, correct = F) # basic chi-square test
model8.2$expected
model8.2
```

Rerun the above but as a GLM (without any random factors). Again, highly significant interaction effect
```{r}
M0.2 <- glm(Strep_outcomeLW ~ 1, family = binomial, data = assay2.2)
M1.2 <- glm(Strep_outcomeLW ~ medium2, family = binomial, data = assay2.2)
summary(M0.2)
summary(M1.2)
anova(M0.2, M1.2, test = "Chi")  
AIC(M0.2, M1.2)

M0.8 <- glm(Strep_outcomeLW ~ 1, family = binomial, data = assay8.2)
M1.8 <- glm(Strep_outcomeLW ~ medium2, family = binomial, data = assay8.2)
summary(M0.8)
summary(M1.8)
anova(M0.8, M1.8, test = "Chi")  
AIC(M0.8, M1.8)
```

However, there are correlated residuals in the model without a random factor, by Softbug_strain
```{r}
E1.2 <- rstandard(M1.2)
E1.8 <- rstandard(M1.8)
par(mfrow=c(2, 1))
plot(E1.2 ~ assay2.2$Softbug_strain); abline(0,0) 
plot(E1.8 ~ assay8.2$Softbug_strain); abline(0,0) 
par(mfrow=c(1, 1))
# rm(M1)
# rm(E1)
```

Now the proper test including a random factor as intercept:  Softbug_strain (i.e. the non-antibiotic-producer strains, of which there are 10)

Highly significant interaction effect.  p = 2.2e-16
```{r}
M2 <- glmer(Strep_outcomeLW ~ 1 + (1|Softbug_strain), family = binomial, data = assay2.2)
M3 <- glmer(Strep_outcomeLW ~ medium2 + (1|Softbug_strain), family = binomial, data = assay2.2)
summary(M3)
anova(M2, M3) # p = 2.2e-16
fixef(M3) # fewer wins (1s) on Control media
plot_model(M3, type = "re")

M4 <- glmer(Strep_outcomeLW ~ 1 + (1|Softbug_strain), family = binomial, data = assay8.2)
M5 <- glmer(Strep_outcomeLW ~ medium2 + (1|Softbug_strain), family = binomial, data = assay8.2)
summary(M5)
anova(M4, M5) # p = 2.2e-16
fixef(M5) # fewer wins (1s) on Control media
plot_model(M5, type = "re")
```

Correlated residuals have largely disappeared
```{r}
par(mfrow=c(2, 1))
plot(assay2.2$Softbug_strain, resid(M2), xlab="Softbug_strain", ylab="Residuals"); abline(0,0)
plot(assay8.2$Softbug_strain, resid(M5), xlab="Softbug_strain", ylab="Residuals"); abline(0,0)
par(mfrow=c(1,1))
```
