---
title: "Figure 1A.  Analyses of Growth rate experiment using Softbugs isolated from the environment"
author: "Douglas Yu"
date: "05 June 2018"
output: html_document
---
## Packages
```{r, message=FALSE}
# rm(list=ls())
library(tidyverse)
library(readxl)
library(knitr)
library(RColorBrewer)
library(ggthemes)
library(lattice)
library(lme4)
library(sjPlot)
library(envDocument)
env_doc("table", git = FALSE)
```

```{r}
datafile <- "Figure_1A_Growth_rate_analyses_DataSheet_20180302.xlsx"
sheet_strepto <- "data_table_Strep"
sheet_softbugs <- "data_table_Sb"
```

## Dataset 1:  _Streptomyces_
```{r Streptomyces dataset}
assay1 <- read_excel(datafile, sheet_strepto, na = "NA")
tbl_df(assay1) # should have 492 rows and 7 cols
assay1 <- assay1 %>% dplyr::select(-Input.order)
assay1 # should have 492 rows and 6 cols
```

## Dataset 2:  Softbugs (our shorthand term for _Non-antibiotic-producers_)
```{r Softbugs dataset}
assay2 <- read_excel(datafile, sheet_softbugs, na = "NA")
tbl_df(assay2) # should have 483 rows and 10 cols
assay2 <- assay2 %>% dplyr::select(-Input.order)
assay2$Inv.strain <- assay2$Inv.strain + 10 # to differentiate numbering from Strepto samples
assay2 <- assay2 %>% mutate(Plate = ifelse(Plate == 4, 1, ifelse(Plate == 5, 2, ifelse(Plate == 6, 3, NA))))
assay2 # should have 483 rows and 6 cols
```

## Assay 3:  Combined 
```{r assay3, message=FALSE, warning=FALSE}
assay3 <- bind_rows(assay1, assay2)
assay3 <- assay3 %>% dplyr::select(Ps.genotype, Invader, Inv.strain, Ps.strain, Plate, Invasion.score) %>% arrange(Ps.genotype, Invader, Inv.strain, Ps.strain, Plate)
assay3 <- assay3 %>% filter(!is.na(Invasion.score))
assay3$Ps.genotype <- as.factor(assay3$Ps.genotype)
assay3$Ps.genotype <- fct_relevel(assay3$Ps.genotype, "Control", "Ps1", "Ps2")
assay3$Inv.strain <- as_factor(as.character(assay3$Inv.strain))
assay3$Inv.strain_shape <- if_else(assay3$Inv.strain == "18", "2", "16") 
assay3$Invader <- as_factor(assay3$Invader) %>% fct_relevel("Soft_bug", "Streptomyces")
assay3$Ps.strain <- as_factor(assay3$Ps.strain)
assay3$Plate <- as_factor(as.character(assay3$Plate))
tbl_df(assay3) # 975 rows and 7 cols
```


Create crosstabs to see experimental design
```{r}
assay4 <- assay3 %>% dplyr::filter(Ps.strain != "Control") 
assay4$Ps.strain <- droplevels(assay4$Ps.strain)
xtabs(~ Inv.strain + Ps.strain + Plate, assay4)
rm(assay4)

assay5 <- assay3 %>% dplyr::filter(Ps.strain == "Control")
assay5$Ps.strain <- droplevels(assay5$Ps.strain)
xtabs(~ Inv.strain + Ps.strain + Plate, assay5)
rm(assay5)
```

sample sizes
```{r}
assay1 %>% count(Invader, Ps.genotype)
assay2 %>% count(Invader, Ps.genotype)
assay3 %>% count(Invader, Ps.genotype, Plate)
```
To read these tables:

1. We infused the media with antibiotics from 17 _Pseudonocardia_ lines that had been collected from different ant colonies (A-S), plus a non-infused control media. There are potentially genetic differences among these vertically transmitted lineages, so we call them:  Ps.strain

2. We measured growth rates of 10 strains of _Streptomyces_ (1-10) and 10 strains of Softbugs (11-20). These are invading strains so we call them:  Inv.strain. 

3. Each of the (17 \* 20 =) 340 combinations of Ps.strain and Inv.strain was repeated 3 times for a total sample size of (340 \* 3 =) 1020 treatment inoculations and 60 control inoculations, but some colonies could not be scored due to difficulties in reading colony area, leaving us with (170 + 292 =) 462 _Streptomyces_ inoculations and (149 + 304 =) 453 Softbugs inoculations. 462 + 453 = 915 total treatment inoculations + (20 \* 3 =) 60 control inoculations = 975 total inoculations.

4. Thus, the experiment is fully crossed but somewhat imbalanced at the end. 

5. (510 - 462 =) 48 Streptomyces inoculations and (510 - 453 =) 57 Softbug inoculations could not be scored. Note that we did score cases where there was no growth. 

```{r stripchart of results}
colorvec <- brewer.pal(3,"RdBu")
colorvec <- c(colorvec[3], colorvec[1], colorvec[2])

p <- ggplot(assay3, aes(x = Ps.genotype:Invader, y = Invasion.score)) 

# truncated y-axis version for publication
(q <- p + ylim(0, 2.5) 
	+ geom_boxplot(size = .5, outlier.shape = NA, width = 0.3, aes(fill = assay3$Invader)) 
	+ geom_jitter(width = 0.05, cex = .8, aes(shape = assay3$Inv.strain_shape)) 
	+ scale_shape_manual(values = c(16, 2)) 
	+ labs(y = expression("Growth score:  Mean colony area " ~ (cm^{2}))) 
	# + scale_fill_brewer(type = "qual", palette = "Set3") # alternative color set
	+ scale_fill_manual(values = colorvec) 
    + theme(axis.title.y = element_text(size = 14), 
  			axis.title.x = element_blank(), 
  			legend.position='none', 
  			panel.background = element_rect(fill = "white"), 
  			panel.grid.major.y = element_line(colour = "grey", size = 0.5),
  			panel.grid.minor.y = element_line(colour = "grey", size = 0.1),
  			panel.border = element_rect(colour = "black", fill = NA)
  			)
	) 
```

```{r}
# non-truncated y-axis version. Analysis uses the whole dataset
q + ylim(0, 4) 
```

To read these boxplots:

1. The y-axis is the final area (cm^2^) of the colony inoculated onto the plate (the notional 'invader')

2. On control media (left 2 columns), Soft-bug colonies (blue) grow bigger (more quickly) than do _Streptomyces_ colonies (orange). 

3. However, on _Pseudonocardia_-infused media (Ps1, Ps2), _Streptomyces_ colonies (orange) grow more quickly than do soft-bug colonies (blue).

4. The 17 _Pseudonocardia_ colonies (Ps.strain) used to infuse the media can be usefully divided into two 16S genotypes (Ps1 n = 6, Ps2 n = 11). Growth rates of _Streptomyces_ and softbugs appear lower on the Ps2 infusions (i.e. Ps2 appears to be a more effective inhibitor in general).

5. There are fewer Control datapoints because this represents only one kind of media, whereas the _Pseudonocardia_ (Ps1 and Ps2) treatment represents 17 Ps lineages. 

6. The empty triangles are all from Softbug strain 18, clearly demonstrating the existence of correlated residuals in the Inv.strain variable. 


Ths dataset is used to choose strains for the pairwise competition experiment (Fig 2)
```{r assay4 summary data tables, eval=FALSE}
(assay4.1 <- assay3 %>% group_by(Ps.genotype, Invader) %>% summarise(median.Invasion.score=round(median(Invasion.score), 2)))

assay4.2 <- assay3 %>% group_by(Ps.genotype, Invader, Inv.strain) %>% summarise(median.Invasion.score=round(median(Invasion.score), 2))
assay4.2 <- assay4.2 %>% arrange(Ps.genotype, Invader, median.Invasion.score)

assay4.3 <- assay3 %>% group_by(Ps.genotype, Ps.strain, Invader, Inv.strain) %>% summarise(median.Invasion.score=round(median(Invasion.score), 2))

# write.table(assay4.1, file = "assay4.1.tsv", quote = FALSE, sep = "\t") # use this to choose summary pairs for the competition experiment
# write.table(assay4.2, file = "assay4.2.tsv", quote = FALSE, sep = "\t") # use this to choose summary pairs for the competition experiment
# write.table(assay4.3, file = "assay4.3.tsv", quote = FALSE, sep = "\t") # use this to choose summary pairs for the competition experiment

rm(assay4.1, assay4.2, assay4.3)
```


## Start of analysis. 

There are two potential random factors:  

1. _Pseudonocardia_ strains: `Ps.strain` (17 levels: A-S) 
2. Invader strains: `Inv.strain` (20 levels: labeled 1-20). 1-10 are _Streptomyces_ strains, 11-20 are Softbug strains

There are 3 Plate replicates for each combination of Ps.strain and Inv.strain, and each of the 20 Inv.strains is tested against each of the 17 Ps.strains. Some complexity:  not all inoculations were scorable, so were omitted. This is a fully crossed but somewhat imbalanced experiment

There are 2 fixed factors:  

1. Pseuonocardia genotype: `Ps.genotype` (Ps1, Ps2).
2. Invader type: `Invader` (softbugs, _Streptomyces_)

Finally, the response variable is:

1. Mean colony area (cm^2^): `Invasion.score`

Construct a full model and test for an interaction effect of medium (Ps.genotype) and Invader (_Streptomyces_ and Softbugs) and also test for correlated residuals in each of the potential random factors

```{r Check for correlated residuals}
M1 <- lm(Invasion.score~Invader*Ps.genotype, data=assay3)
E1 <- rstandard(M1)
plot(E1 ~ assay3$Ps.strain); abline(0,0) # T,P,Q strains either mostly above or below 0, demonstrating correlated residuals for some strains
plot(E1 ~ assay3$Inv.strain); abline(0,0) # 11,14,16,17,18,19,20,4,5,9,10 either mostly above or below 0, demonstrating correlated residuals for some strains. recall that 11-20 are softbugs and 1-10 are _Streptomyces_
rm(M1)
rm(E1)
```

Thus, we use a model with random factors. In particular, note that Softbug strain 18 has a high growth rate, which we saw in the boxplot (empty triangle points)

```{r}
# with interaction effect
M1 <- lmer(Invasion.score ~ Invader * Ps.genotype + (1|Ps.strain/Plate) + (1|Inv.strain), data=assay3, REML = FALSE)
# without interaction effect
M2 <- lmer(Invasion.score ~ Invader + Ps.genotype + (1|Ps.strain/Plate) + (1|Inv.strain), data=assay3, REML = FALSE)

anova(M1, M2) # highly significant interaction effect:  p = 1.102e-10
AIC(M1, M2) # AIC difference >> 2, and the model with the interaction effect has a much lower AIC:  263.39
```

Estimate parameter values
```{r}
M3 <- lmer(Invasion.score ~ Invader * Ps.genotype + (1|Ps.strain/Plate) + (1|Inv.strain), data=assay3, REML = TRUE)
summary(M3)
```

There is a highly significant interaction effect. Softbugs grow more quickly on the Control medium, but Streptomyces grow more quickly on both the Ps1- and Ps2-infused medias.

Random factors:
```{r}
plot(assay3$Ps.strain, resid(M2), xlab="Ps.strain", ylab="Residuals"); abline(0,0)
plot(assay3$Inv.strain, resid(M2), xlab="Inv.strain", ylab="Residuals"); abline(0,0)
plot(assay3$Plate, resid(M2), xlab="Inv.strain", ylab="Residuals"); abline(0,0)
```

The corelated residuals have gone away, as expected, although Inv.strain 18 has greater variance than the other strains. 

### Effect of Ps1 vs Ps2

From the Andersen et al. (2013) results, we hypothesised that Ps2 would be a stronger growth suppressor than is Ps1.

Remove the Control level and test again for a significant main effect of Ps.genotype (Ps1 vs Ps2).
```{r}
assay4 <- assay3 %>% dplyr::filter(Ps.genotype != "Control")
assay4$Ps.genotype <- fct_drop(assay4$Ps.genotype)

M5 <- lmer(Invasion.score ~ Invader + Ps.genotype + (1|Ps.strain/Plate) + (1|Inv.strain), data=assay4, REML = FALSE)
M6 <- lmer(Invasion.score ~ Invader + (1|Ps.strain) + (1|Inv.strain/Plate), data=assay4, REML = FALSE)

anova(M5, M6)
```


Ps.genoptype is highly significant (p = 7.252e-07):  growth is different between the Ps1 and Ps2-infused media

Now estimate parameter value
```{r}
M7 <- lmer(Invasion.score ~ Invader + Ps.genotype + (1|Ps.strain/Plate) + (1|Inv.strain), data=assay4, REML = TRUE)
summary(M7)
```

Growth rate is lower on Ps2 (-0.16489)





