---
title: "Figure 3  Analyses of MIC and LEC scores"
author: "Douglas Yu"
date: "05 June 2018"
output: html_document
---
## Packages
```{r, message=FALSE}
# rm(list=ls())
library(tidyverse)
library(readxl)
library(cowplot)
library(knitr)
library(ggbeeswarm)
library(RColorBrewer)
library(ggthemes)
library(envDocument)
env_doc("table", git = FALSE)
```

```{r}
datafile <- "Figure_3_MIC_LEC_analyses_DataSheet_20180531.xlsx"
datasheet <- "Sheet1"
```

## Dataset 1: MIC and LEC scores
```{r Streptomyces dataset}
assay1 <- read_excel(datafile, datasheet, na = "NA")
assay1 <- assay1 %>% dplyr::select(Strain, Strep_Soft, Antibiotic, MIC_score, LEC_score, Notes)
assay1 <- assay1 %>% dplyr::filter(is.na(Notes)) %>% dplyr::select(-Notes)
assay1$Strain <- as_factor(assay1$Strain)
assay1$Strep_Soft <- as_factor(assay1$Strep_Soft)
assay1$Antibiotic <- as_factor(assay1$Antibiotic)
tbl_df(assay1) # should have 155 rows and 5 cols

xtabs(~ Strain + Strep_Soft, assay1) # to see experimental design
```


```{r stripchart of results}
colorvec <- brewer.pal(3,"RdBu")
colorvec <- c(colorvec[3], colorvec[1], colorvec[2])

mic <- ggplot(assay1, aes(x = Strep_Soft, y = MIC_score))
lec <- ggplot(assay1, aes(x = Strep_Soft, y = LEC_score)) 

(micplot <- mic 
	+ geom_boxplot(size = .5, outlier.shape = NA, width = 0.3, aes(fill = assay1$Strep_Soft), notch = TRUE) 
	# + geom_jitter(width = 0.05, cex = .8)
    + geom_quasirandom()
	+ scale_shape_manual(values = c(16, 2)) 
	+ labs(y = expression("Minimum Inhibitory Concentration (MIC) score")) 
	+ scale_fill_manual(values = colorvec) 
    + theme(axis.title.y = element_text(size = 14), 
  			axis.title.x = element_blank(), 
  			legend.position='none', 
  			panel.background = element_rect(fill = "white"), 
  			panel.grid.major.y = element_line(colour = "grey", size = 0.5),
  			panel.grid.minor.y = element_line(colour = "grey", size = 0.1),
  			panel.border = element_rect(colour = "black", fill = NA)
  			)
    )

(lecplot <- lec 
	+ geom_boxplot(size = .5, outlier.shape = NA, width = 0.3, aes(fill = assay1$Strep_Soft), notch = TRUE) 
	# + geom_jitter(width = 0.05, cex = .8)
    + geom_quasirandom()
	+ scale_shape_manual(values = c(16, 2)) 
	+ labs(y = expression("Lowest Inhibitory Concentration (LEC) score")) 
	+ scale_fill_manual(values = colorvec) 
    + theme(axis.title.y = element_text(size = 14), 
  			axis.title.x = element_blank(), 
  			legend.position='none', 
  			panel.background = element_rect(fill = "white"), 
  			panel.grid.major.y = element_line(colour = "grey", size = 0.5),
  			panel.grid.minor.y = element_line(colour = "grey", size = 0.1),
  			panel.border = element_rect(colour = "black", fill = NA)
  			)
    )

plot_grid(micplot, lecplot, labels = c("A", "B"))
```

```{r statistics}
median(assay1$MIC_score[assay1$Strep_Soft == "Antibiotic-producers"])
median(assay1$MIC_score[assay1$Strep_Soft == "Non-producers"])

median(assay1$LEC_score[assay1$Strep_Soft == "Antibiotic-producers"])
median(assay1$LEC_score[assay1$Strep_Soft == "Non-producers"])

wilcox.test(assay1$MIC_score ~ assay1$Strep_Soft)
wilcox.test(assay1$LEC_score ~ assay1$Strep_Soft)
```

```{r summarise by Strain}
assay2 <- assay1 %>% group_by(Strain) %>% 
    summarise(Strep_Soft = first(Strep_Soft), 
                 mean_MIC = mean(MIC_score),
                 mean_LEC = mean(LEC_score)
    )

wilcox.test(assay2$mean_MIC ~ assay2$Strep_Soft)
wilcox.test(assay2$mean_LEC ~ assay2$Strep_Soft)
```

To read these results:

1. Antibiotic-producers have higher MIC and LEC scores.

2. We conservatively take the mean MIC and LEC scores (across all antibiotics) for each Antibiotic-producer and each Non-producer strain, reducing the dataset from n = 155 to n = 20:  assay2

3.  We then apply a Wilcoxon 2-sided test of equality of means.  Mean MIC scores and LEC scores are significantly higher for Antibiotic-producers.  

